Imputation of incomplete multilevel data

### 2024-06-06 manuscript.qmd
re-order imputation workflow

@@

1. missing data occur often in data with human subjects
2. missing data may be resolved, but need to be handled in accordance with the analysis of scientific interest
3. in human-subjects research, there is often clustering, which may be captured with multilevel modeling techniques
4. if the analysis of scientific interest is a multilevel model, the missing data handling method should accommodate the multilevel structure of the data
4. both missingness and multilevel structures require advanced statistical techniques
5. this tutorial sets out to facilitate empirical researchers in accommodating both multilevel structures as well as missing data
6. we illustrate the use of the software by means of a case study

@@

FIMD, section 7.10

> Recipe for a level-1 target:

> 1.	Define the most general analytic model to be applied to imputed data;
2.	Select a 2l method that imputes close to the data;
3.	Include all level-1 variables;
4.	Include the disaggregated cluster means of all level-1 variables;
5.	Include all level-1 interactions implied by the analytic model;
6.	Include all level-2 predictors;
7.	Include all level-2 interactions implied by the analytic model;
8.	Include all cross-level interactions implied by the analytic model;
9.	Include predictors related to the missingness and the target;
10.	Exclude any terms involving the target

> Recipe for a level-2 target:

> 1.	Define the most general analytic model to be applied to imputed
data;
2.	Select a 2lonly method that imputes close to the data;
3.	Include the cluster means of all level-1 variables;
4.	Include the cluster means of all level-1 interactions;
5.	Include all level-2 predictors;
6.	Include all interactions of level-2 variables;
7.	Include predictors related to the missingness and the target;
8.	Exclude any terms involving the target

@@

1. Load the data, make sure the variables are correctly formatted (e.g. numeric clustering variable)
2. Explore the missingness
2. ~~Fix any cluster-level missingness determinately (note that `2l.only` methods do not work if there are inconsistencies, fix those first!). Re-evaluate the missingness~~
3. Explore bivariate relations and associations with missingness indicators (optionally test the associations?)
4. For each incomplete variable:
  - Determine the imputation method
  - Choose imputation model predictors (see recipe)
5. Set methods vector and predictor matrix
6. Impute the incomplete data
6. If error...
6. After successful run:
  - Evaluate convergence
  - Evaluate imputations
9. Analyze and pool
9. If error... (e.g. rescale variables by converting imputations to long format, mutate, then back to mids)
9. After successful run:
  - Evaluate missingness ratio's (e.g. fraction of missing information)
  - If necessary, increase $m$ and re-impute
10. For further multilevel model building, AICs could be pooled, how?

@@

Load the data
```{r}
load("data/popular_MAR.RData")
dat <- popular_MAR
```

Inspect the missingness.
```{r}
plot_pattern(dat, square = FALSE)
```

@@

```{r}
plot_corr(dat, square = FALSE)
plot_flux(dat, label = FALSE)
```

@@

## Complete case analysis

Before we start imputing the incomplete data, we can fit a complete case model to the observed data. This will give us a baseline for comparison with the imputed data.

Single-level model
```{r}
fit <- lm(
  popularity_ij ~ 1, 
  data = dat
  )
tidy(fit)
```

M0: intercept-only model
```{r}
fit <- lmer(
  popularity_ij ~ (1 | cluster_id), 
  data = dat, 
  REML = FALSE
  )
tidy(fit)
```

M1: model with explanatory variables
```{r}
fit <- lmer(
  popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1 | cluster_id), 
  data = dat, 
  REML = FALSE
  )
tidy(fit)
```

M1A: model with explanatory variables, extraversion slope random
```{r}
fit <- lmer(
  popularity_ij ~ gender_ij + extraversion_ij + experience_j + 
    (1  + extraversion_ij | cluster_id), 
  data = dat, 
  REML = FALSE
  )
tidy(fit)
```

M0: model with with cross-level interaction
```{r}
fit <- lmer(
  popularity_ij ~ gender_ij + extraversion_ij + experience_j + 
    extraversion_ij:experience_j + (1  + extraversion_ij | cluster_id), 
  data = dat, 
  REML = FALSE
  )
tidy(fit)
```

The last model does not converge. This is likely due to the missingness in the data. We will now impute the incomplete data and re-fit the models.

## Imputation model building

@@

```{r}
ggmice(dat, aes(popularity_ij)) + 
  geom_histogram(fill = "white")
```

@@

We can use `2l.norm`: a multilevel method using the linear mixed model with heterogeneous error variances.

@@

@@

@@

@@

Our final predictor matrix can be visualized as follows.
```{r}
plot_pred(pred, meth = meth)
```

Impute the data
```{r, eval=FALSE}
imp <- mice(
  dat, 
  m = 1, 
  method = meth, 
  predictorMatrix = pred,
  maxit = 1)
```

@@

This code runs! Add iterations and evaluate convergence.
```{r}
imp <- mice.mids(imp, maxit = 9, printFlag = FALSE)
plot_trace(imp)
```

@@

After 20 iterations, there is no apparent trending visible in the trace plots. We will first evaluate the imputations of the variables with atypical trace plots.

@@

# Visualize

```{r}
ggmice(imp, aes(y = popularity_ij, x = extraversion_ij, group = cluster_id)) +
  geom_point(alpha = 0.1) +
  geom_smooth(color = "grey", method = "lm", se = FALSE, alpha = 0.1) +
  geom_smooth(aes(group = NULL), color = "black", method = "lm", se = FALSE, alpha = 0.1)
```

### Analysis

@@

```{r, echo=TRUE}
cca <- lm(
  popularity_ij ~ 1, 
  data = dat
  )
tidy(cca)
```

```{r, echo=TRUE}
fit <- with(
  imp,
  lm(popularity_ij ~ 1, data = dat)
  ) |> 
  pool()
tidy(fit)
```

>>>>

```{r}
library(knitr)
# print everything as kable ----
knit_print.data.frame <- function (x, options, ...) {
  knitr::kable(x) |> knitr::knit_print(options, ...)
}
registerS3method("knit_print", "data.frame", knit_print.data.frame)
options(scipen = 999, digits = 3)
```

@@

<!-- 1. missing data occur often in data with human subjects -->
<!-- 2. missing data may be resolved, but need to be handled in accordance with the analysis of scientific interest -->
<!-- 3. in human-subjects research, there is often clustering, which may be captured with multilevel modeling techniques -->
<!-- 4. if the analysis of scientific interest is a multilevel model, the missing data handling method should accommodate the multilevel structure of the data -->
<!-- 4. both missingness and multilevel structures require advanced statistical techniques -->
<!-- 5. this tutorial sets out to facilitate empirical researchers in accommodating both multilevel structures as well as missing data -->
<!-- 6. we illustrate the use of the software by means of a case study -->

@@

<!-- FIMD, section 7.10 -->

<!-- > Recipe for a level-1 target: -->

<!-- > 1.	Define the most general analytic model to be applied to imputed data; -->
<!-- 2.	Select a 2l method that imputes close to the data; -->
<!-- 3.	Include all level-1 variables; -->
<!-- 4.	Include the disaggregated cluster means of all level-1 variables; -->
<!-- 5.	Include all level-1 interactions implied by the analytic model; -->
<!-- 6.	Include all level-2 predictors; -->
<!-- 7.	Include all level-2 interactions implied by the analytic model; -->
<!-- 8.	Include all cross-level interactions implied by the analytic model; -->
<!-- 9.	Include predictors related to the missingness and the target; -->
<!-- 10.	Exclude any terms involving the target -->

<!-- > Recipe for a level-2 target: -->

<!-- > 1.	Define the most general analytic model to be applied to imputed -->
<!-- data; -->
<!-- 2.	Select a 2lonly method that imputes close to the data; -->
<!-- 3.	Include the cluster means of all level-1 variables; -->
<!-- 4.	Include the cluster means of all level-1 interactions; -->
<!-- 5.	Include all level-2 predictors; -->
<!-- 6.	Include all interactions of level-2 variables; -->
<!-- 7.	Include predictors related to the missingness and the target; -->
<!-- 8.	Exclude any terms involving the target -->

@@

<!-- 1. Load the data, make sure the variables are correctly formatted (e.g. numeric clustering variable) -->
<!-- 2. Explore the missingness -->
<!-- 2. ~~Fix any cluster-level missingness determinately (note that `2l.only` methods do not work if there are inconsistencies, fix those first!). Re-evaluate the missingness~~ -->
<!-- 3. Explore bivariate relations and associations with missingness indicators (optionally test the associations?) -->
<!-- 4. For each incomplete variable: -->
<!--   - Determine the imputation method -->
<!--   - Choose imputation model predictors (see recipe) -->
<!-- 5. Set methods vector and predictor matrix -->
<!-- 6. Impute the incomplete data -->
<!-- 6. If error... -->
<!-- 6. After successful run: -->
<!--   - Evaluate convergence -->
<!--   - Evaluate imputations -->
<!-- 9. Analyze and pool -->
<!-- 9. If error... (e.g. rescale variables by converting imputations to long format, mutate, then back to mids) -->
<!-- 9. After successful run: -->
<!--   - Evaluate missingness ratio's (e.g. fraction of missing information) -->
<!--   - If necessary, increase $m$ and re-impute -->
<!-- 10. For further multilevel model building, AICs could be pooled, how? -->

@@

Load the complete data.
```{r}
load("data/popular.RData")
```

Load the incomplete data.
```{r}
load("data/popular_MAR.RData")
dat <- popular_MAR
```

## Comparative truth

Sequentially fit multilevel models to the complete data to obtain the true parameter estimates.

```{r}
mod_true_singlelevel <- lm(popularity_ij ~ 1, data = popular) 
```

```{r}
mod_true_singlelevel |> tidy()
```

```{r}
mod_true_intercept <- lmer(popularity_ij ~ (1 | cluster_id), data = popular, REML = FALSE) 
```

```{r}
mod_true_intercept |> tidy()
```

```{r}
mod_true_predictors <- lmer(popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1 | cluster_id), data = popular, REML = FALSE)
```

```{r}
mod_true_predictors |> tidy()
```

```{r}
mod_true_randomslope <- lmer(popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1  + extraversion_ij | cluster_id), data = popular, REML = FALSE)
```

```{r}
mod_true_randomslope |> tidy()
```

```{r}
mod_true_interaction <- lmer(popularity_ij ~ gender_ij + extraversion_ij + experience_j + extraversion_ij:experience_j + (1  + extraversion_ij | cluster_id), data = popular, REML = FALSE)
```

```{r}
mod_true_interaction |> tidy()
```

## Complete-case analysis

Sequentially fit multilevel models to the incomplete data to obtain the parameter estimates after list-wise deletion.

```{r}
mod_cca_singlelevel <- lm(popularity_ij ~ 1, data = dat) 
```

```{r}
mod_cca_singlelevel |> tidy()
```

```{r}
mod_cca_intercept <- lmer(popularity_ij ~ (1 | cluster_id), data = dat, REML = FALSE) 
```

```{r}
mod_cca_intercept |> tidy()
```

```{r}
mod_cca_predictors <- lmer(popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1 | cluster_id), data = dat, REML = FALSE)
```

```{r}
mod_cca_predictors |> tidy()
```

```{r}
mod_cca_randomslope <- lmer(popularity_ij ~ gender_ij + extraversion_ij + experience_j + (1  + extraversion_ij | cluster_id), data = dat, REML = FALSE)
```

```{r}
mod_cca_randomslope |> tidy()
```

```{r}
mod_cca_interaction <- lmer(popularity_ij ~ gender_ij + extraversion_ij + experience_j + extraversion_ij:experience_j + (1  + extraversion_ij | cluster_id), data = dat, REML = FALSE)
```

```{r}
mod_cca_interaction |> tidy()
```

<!-- 1.	Inspect incomplete data -->
<!-- a.	Explore variables: determine imputation methods with str() and ggmice(),  -->
<!-- b.	If necessary, add variables for polynomial terms -->
<!-- c.	Explore missingness: evaluate missing data patterns with plot_pattern() -->
<!-- d.	Determine imputation model predictors using plot_corr(), plot_flux(), and ggmice() -->
<!-- 2.	Build imputation models -->
<!-- a.	Create methods vector with imputation methods using make.methods() -->
<!-- b.	Create predictor matrix with imputation model predictors using make.predictorMatrix() -->
<!-- c.	Evaluate imputation models using plot_pred() -->
<!-- 3.	Run imputation algorithm -->
<!-- a.	Impute: fill in the missing values with mice() -->
<!-- b.	Inspect convergence: visualize algorithmic convergence with plot_trace() -->
<!-- c.	If necessary, add iterations -->
<!-- 4.	Evaluate imputations -->
<!-- a.	Inspect imputed values: assess imputed values with ggmice(), PPC? -->
<!-- b.	Analyze imputed data: fit the model of scientific interest with lmer()   -->
<!-- c.	Evaluate missingness metrics with testEstimates() -->
<!-- d.	If necessary, add imputations -->
<!-- 5.	Interpret results and report -->

## Inspect incomplete data

### Explore observed data

Assess the structure of the data.
```{r}
str(dat)
```

Print the first few rows of the data.
```{r}
head(dat)
```

### Explore missingness

Inspect the missing data patterns.
```{r}
plot_pattern(dat, square = FALSE)
```

@@

```{r}
plot_corr(dat, square = FALSE)
```

@@

```{r}
plot_flux(dat) 
```

## Build imputation models

@@

```{r}
ggmice(dat, aes(popularity_ij)) + 
  geom_density()
```

@@

Further inspect the distribution of the outcome variable in the first few clusters.
```{r}
dat |> 
  filter(cluster_id %in% unique(dat$cluster_id)[1:4]) |>
  ggmice(aes(popularity_ij)) + 
    geom_histogram(fill = "white") +
    facet_wrap(~cluster_id)
```

We will use `2l.norm`: a multilevel method using the linear mixed model with heterogeneous error variances.

@@

#### Teacher experience

@@

#### Methods vector

@@

#### Predictor matrix

@@

Our final predictor matrix is as follows.
```{r}
pred
```

### Evaluate imputation models

The imputation models can be visualized as follows.
```{r}
plot_pred(pred, meth = meth)
```

## Run imputation algorithm

### Impute

Impute the data using the specified imputation methods and predictor matrix.

```{r, eval=FALSE}
imp <- mice(
  dat, 
  m = 1, 
  method = meth, 
  predictorMatrix = pred,
  maxit = 1)
```

@@

This code runs! Add iterations to reach algorithmic convergence.
```{r}
imp <- mice.mids(imp, maxit = 9, printFlag = FALSE)
```

### Evaluate convergence

Check for algorithmic non-convergence.

```{r}
plot_trace(imp)
```

@@

After 20 iterations, there is no apparent trending visible in the trace plots.

## Evaluate imputations

### Inspect imputed values

We will first evaluate the imputations of the variables with atypical trace plots.

@@

<!-- # Visualize -->

<!-- ```{r} -->
<!-- ggmice(imp, aes(y = popularity_ij, x = extraversion_ij, group = cluster_id)) + -->
<!--   geom_point(alpha = 0.1) + -->
<!--   geom_smooth(color = "grey", method = "lm", se = FALSE, alpha = 0.1) + -->
<!--   geom_smooth(aes(group = NULL), color = "black", method = "lm", se = FALSE, alpha = 0.1) -->
<!-- ``` -->

### Analyze imputed data

@@

```{r, echo=TRUE}
mod_cca_singlelevel <- lm(
  popularity_ij ~ 1, 
  data = dat
  )
tidy(cca)
```

```{r, echo=TRUE}
mod_imp_singlelevel <- with(
  imp,
  lm(popularity_ij ~ 1, data = dat)
  ) |> 
  pool()
tidy(fit)
```
